Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-16 Reviews & Rebuttal.md

Rebuttal & Coverletter
======================

This rebuttal/coverletter is written in markdown. A graphical summary of the reviews can be found [here](https://docs.google.com/spreadsheets/d/1DTPt5eGe6PpMiNh38p4YsQrKLI3bqKhylZQKZKiZg4w/edit?usp=sharing).

General Changes
---------------

- substantial reductions had to take place, since we were given a `short paper` slot with an 8-page limit.
	- redundancy between `Introduction` and `Epidemiological Background` was minimized as much as possible
	- `Related Work` section was shortened and covers less papers to meet the page limit.
		- as requested by [Reviewer 3], we elaborate on the differences towards the VMV paper in the `Related Work` section
	- `The Lumbar Spine Data Set` section was shortened, `Figure 2` was removed since we found it did not support the train of thought as we intended
	- `Experiments and Preliminary Results` section was shortened and changed due to the Reviewer feedback. It should be more precise and comprehensible.
- `Results` section now covers *all subjects*, subjects categorized by *gender* and *body mass index*
	- `Figure 3` (former `Figure 4`) was adapted accordingly
- `Summary & Conclusion` was completely revised and is now structured properly
	- due to the lack of time until the camera ready deadline as well as the page limit, we are not able to conduct a detailed user test. We did, however, discuss our approach and its drawbacks in more detail in the last section
- Figure 1 was completely removed since it does not supports the train of thought as we intended and due to lack of space
- We named our decision tree visualization *decision tree quality plot*

### 2014-12-17 draft.tex
removed comments from the LaTeX source
fixed incorrect subject count for image detection

% modify floating spacings 
% \addtolength{\abovecaptionskip}{-0.0mm}
% \addtolength{\belowcaptionskip}{-0.0mm}
% \addtolength{\dbltextfloatsep}{-5.0mm}
% \addtolength{\textfloatsep}{-4mm}
% \addtolength{\intextsep}{-5.0mm}
%
% % modify paragraph spacings
% \addtolength{\parskip}{-5.0mm}

@@

% \author{Blind}
\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1}, Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{klemm@isg.cs.uni-magdeburg.de}
}

@@

\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as spine canal curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
%%
Therefore, we automatically train a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with a \emph{decision tree quality plot}. %scatter plot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient to describe lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, relations between non-image variables were successfully detected and described.
}

@@

\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
Large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
%%
This information is used to determine risk factors for diseases, inform people about healthier lifestyles or to support the diagnosis of widespread diseases.
%%
% So far, epidemiological research is strongly hypothesis-driven.
% %%
% More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
% %%
% As a result, the hypothesis will be accepted or rejected.
% %%
%%
Cohort studies are an instrument of epidemiological research.
%%
%Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets. %, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
We analyze back pain, one of the most frequent complaints in the Western civilization. %\cite{Hoy2010}. % and gains importance due to the population aging \cite{Hoy2010}.
%%
%Epidemiological analysis of back pain aims to determine risk factors.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar spine attributes is still missing.
%%
%Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
%%

We present an analysis of the image-derived data from a cohort study lumbar spine dataset.
We extract possible associations between spine shape and back pain characteristics.
For this purpose, we combine classification algorithms with data visualization techniques.
Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
% The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
We focus on highlighting new correlations and trigger \emph{hypotheses generation}, rather than statistically validate complex epidemiological correlations.
Our contributions are:
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
\item The identification of lumbar spine shape properties potentially associated with back pain,
\item The detection of associations between image-based, socio-demographic and medical variables for \emph{hypotheses generation},
\item The identification of the most important variables via classification methods using a novel \emph{decision tree quality plot}.
%The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{ivapp15.dnsalias.com}).
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
(1) physicians formulate hypotheses based on observations made in their clinical practice; (2) epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated condition.
% \begin{enumerate}
% %%
% 	\item Physicians formulate hypotheses based on observations made in their clinical practice.
% %%
% 	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
% %%
% 	\item Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
% %%
% \end{enumerate}
Mutually dependent variables make this analysis challenging.
For example, the incidence rate of many diseases, such as different cancer types increases with age \cite{Fletcher}.
%Many diseases, such as different cancer types, are for example more likely with increasing age \cite{Fletcher}.
% For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
% %%
% Age acts as a \emph{risk factor} for prostate cancer.
% %%
% Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\paragraph{Challenges of Epidemiological Data Analysis.}
% Epidemiological data originates from a wide range of studies.
% %%
% The study type depends on the condition of interest.
% %%
% Most common are case-control studies analyzing one specific disease and its influence on the human body.
We focus on epidemiological large scale cohort study data.
% These studies aim at collecting as much data as possible for each subject.
The collected data can be analyzed regarding many diseases and conditions.
Epidemiological data are heterogenous and incomplete.
For example, data about a disease treatment only affects subjects suffering from this condition.
Statistical analysis has to take missing data into account.
Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self-reported questionnaires or genetic examinations.
This yields a heterogenous information space.
Information reduction techniques are necessary to compare these data.
For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.

@@

%Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%These numeric values can be compared with non-image variables to retrieve correlations.

@@

% Epidemiologists assume correlations between lumbar back pain and lifestyle factors.
% %%
% These include \emph{nutrition}, \emph{physical activity} and \emph{body posture combined with physical stress at work}.

@@

%The workflow consists of an iterative sequence of \emph{group selection}, \emph{variable selection} and \emph{visualization}.

@@

%We incorporate the IVA approach by concurrently displaying complex relationships between variables via decision trees.
% \cite{Steenwijk} propose a coordinated linked view system for both image-related and non-image data.
% %%
% It incorporates multiple plot types, such as scatter plots, parallel coordinates and time plots. %with brushing and linking facilities.
% %%
% Similar to our work, they quantify image data and project it into the non-image data information space.

@@

%The extracted lumbar spine shape was correlated with subject size.
%We use the same detection algorithm and lumbar spine canal extraction as foundation of this work.

@@

%Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.

@@

% \cite{Dai} define a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
% %%
% The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.

@@

% Related work in this field often focuses on clinical diagnosis.

@@

% The effort of analyzing decision trees increases with their complexity.

@@

%They present an interactive data mining tool, that can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.

@@

% %%
% We aim at deriving new relationships by training decision trees to explain target variables.
%\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
% \cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
% %%
% The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
% %%
% The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.

@@

% \noindent Our approach allows the simultaneous analysis of many variables.
%Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
\noindent Our work is based on a data set compiled by Epidemiologists with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
%\noindent Epidemiologists compiled a data set with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 6,753 subjects.
%The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.

@@

	% %%
% 	\begin{itemize}
% 		%%
% 		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
% 		%%
% 		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
% 		%%
% 	\end{itemize}

@@

All categorical variables are converted into binary \emph{dummy variables}, indicating the presence or absence of a categorical variable manifestation. %with multiple manifestations into several dichotomous variables.
%Each dichotomous variable encodes the presence of a manifestation.

@@

%One subject can only have one of these variables set to true.
%This allows to determine which manifestation of a variable can be described best using the image data variables.
%This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
% Many variables comprise missing data.
% %%
% Some variables are follow-up questions, covering the treatment of a disease.
% %%
% Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
% %%
% Sparse variables are statistically less resilient and need to be treated with care during the analysis.

@@

%	(a) right: FEM aligned to an MRI data set.
	%The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.

@@

%	We extract the variables for each projection axis to assess their information gain.

@@

% Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
% %%
% The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
% %%

@@

%The tetrahedron-based finite element model (Fig.~\ref{fig:centerline}~a) is initialized using three user-defined landmarks.

@@

%A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describe an initial height estimation (Fig.~\ref{fig:centerline}~a).
%The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
%Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, rotation and scaling.
The detection fails for several subjects due to imaging artifacts or strongly deformed spines, yielding models for $2,540$ out of $6,753$ subjects.
%$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
%We have to assess the model accuracy to extract key figures from it.

@@

%It captures reliable information about spine canal curvature.

@@

% These metrics are also extracted in the sagittal, coronal and transversal projection of the model.
% %%
% We assess the information gain of each dimension using these projections.
% %%
% This yields a total of 9 image-derived variables.

@@

% \begin{figure}[!h]
%   %\vspace{-0.2cm}
%   \centering
%   \includegraphics[width=0.475\textwidth]{figures/age-gender}
%   \caption{
% 	%%
% 	Correlation of \emph{age} and \emph{gender} regarding the lumbar spine size.
% 	%%
% 	The bar chart shows subjects divided into four different \emph{age} groups (x-axis) and their subject count (y-axis).
% 	%%
% 	Each bar contains the mean lumbar spine of the respective group.
% 	%%
% 	The shape color encodes the distance (red for x axis, blue for y axis, green for z axis) to the overall male (top chart) or female (bottom chart) mean shape.
% 	%%
% 	The dark gray share of each bar encodes the portion of male (top chart) or female (bottom chart) subjects.
% 	%%
% %	Women have a higher live expectance than men, hence their higher share in the old age group.
% 	%%
% %	Also, women are on average smaller than men, hence the larger shape similarity with older subjects.
% 	%%
% 	%Due to bone erosion, older people are also smaller on average.
% 	}
%   \label{fig:age-gender}
% \end{figure}
%The first step is to analyze if the dichotomous back pain variable can be discriminated using single shape variables or sets of shape variables.

@@

%In this section, we aim to retrieve correlations between the dichotomous \emph{back pain} indicator and the image-derived variables using a \texttt{GPLOM} as well as a assessment, which non-image variables 

@@

%Since men are on average taller than women and people of old age shrink due to \emph{bone erosion}, \emph{gender} and \emph{age} are also risk factors.
%Women have a higher life expectancy than men, hence their higher share in the old age group.
%Also, women are on average smaller than men, hence the larger shape similarity with older subjects.

@@

%Since the \emph{gender} encodes \emph{body size}, we decided to divide subjects into \emph{body size} groups.

@@

%Since our image-derived variables are metric, their pairwise combinations are visualized using scatter plots.

@@

%The combination of the image variables with back pain is visualized as histogram at the left side of the matrix diagonal and as box plot on the right side.

@@

% \\\\
% \noindent \textbf{Correlation Matrix.}

@@

% We calculated a correlation matrix to assess correlations between the image variables.
% %%
% The pearson correlation coefficients between the numeric variables are depicted right of the matrix diagonal (Fig.~\ref{fig:image-parameter-range}).
% %%
% \emph{Curvature}, \emph{curvature angle} and \emph{torsion} strongly correlate with their planar projections.
% %%
% Also the \emph{mean curvature} and \emph{curvature angle} correlate by a factor of $-0.89$.
% %%
% \emph{Torsion} does not correlate with any other image-derived variable.
%%--------------------------------------------------------
% \\\\
% \noindent \textbf{Correlation of Image Variables With Lumbar Back Pain.}
% Figure~\ref{fig:image-parameter-range} shows the distribution of all image-variables as Generalized Pairs Plot.
% %%
% No statistically significant correlation could be observed through all subject groups.
% %%
% The box plots show no difference between subjects with or without back pain.
% \\\\
% \noindent \textbf{Assessing the Information Gain Using the Principal Component Analysis.}
% A Principal Component Analysis (PCA) of the image data yielded that \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle} explain almost 50\% of the image variance.
% %%
% \emph{Torsion} and metrics of the \emph{transverse} projection planes contribute very little variance, supporting our conclusion about their low information gain.

@@

% To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
% %%
% The PCA results in a linear projection of the variable space ordering their components by maximum variance.
% %%
% The first three principal components explain 75\% of the variance in the nine image-derived variables.
% %%
% The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
% %%
% The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
% %%
% The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
% %%
% This supports our prior conclusion about the low information gain of the transverse planes.
% %%
% \emph{Torsion} also adds little variance to the information space.
% %%
% The complete analysis can be found in the experiments section of the supplementary material (\url{ivapp15.dnsalias.com})
%% ------------------------------------------

@@

%The method uses the following correlation metrics for the different type combinations:
% \begin{itemize}
% %%
% \item \emph{Pearson product-moment} for two continuous variables,
% \item \emph{Polyserial} correlation for one continuous and one categorical variable, and
% \item \emph{Polychoric} correlations for two categorical variables.
% \end{itemize}

@@

These observations brought us to the decision to incorporate classification techniques to assess the influence of the image-derived variables.
%%
% \section{\uppercase{Evaluation of Decision Trees}}
%\section{\uppercase{Interactive Decision Tree Visualization}}
\section{\uppercase{Interactive Decision Tree Quality Plot}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
%%
%The presented method aims to support an explorative analysis by trying to identify relationships with all non-image variables and also 
%The nature of this analysis is more explorative as it includes all 
%%
Decision trees are used to create predictive models.
%%
These models are built w.r.t. all input variables and capture more complex relationships than correlation coefficients.
%%
Leafs of a decision tree represent class labels, branches represent variable conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
%They work with numerical as well as categorical data.
%%
%Readability is only granted for small trees.
%%
%Large, complex trees are difficult to trace and to understand.
%%
Too many branches impose overfitting to the data \cite{DecisionTree}.
%%
\\\\
\noindent \textbf{Creating Decision Trees.}
We use the C4.5 algorithm, which builds decision trees based on information entropy.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%
%\subsection{The C4.5/C5.0 Algorithm}
%\label{sec:C5.0}
%The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
%Such a calculation requires a numeric or categorical target variable.
%%
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  }
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
% %%
% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
% %%
% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%\subsection{Visualizing Decision Trees}

@@

%This is included in our IVA framework.

@@

%Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%The metric usually works well for variables with uniform distribution.
%It distorts the result for other distribution types.

@@

%The error is scaled to denote perfect classification with 0; 1 equals always wrong; 0.5 equals random selection.

@@

%We consider a result rated  or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.

@@

%We still want to be able to display \emph{details-on-demand} and examine a decision tree in detail.

@@

%Selecting a variable using a drop-down menu allows the user to select the variable used for subject subdivisions, e.g. \emph{gender} or \emph{employment status}.

@@

% As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization variables using user interface controls, such as sliders.
% %%
% In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to create reports by combining \texttt{R} with the \texttt{Markdown} syntax.
% %%
% We used \texttt{R Shiny}\footnote{Developed by RStudio, Inc; \texttt{shiny.rstudio.com}} to make the report available as dynamic web application.
% %%
% It allows to combine the power of static \texttt{R Markdown} reports with dynamically parameterized \texttt{ggvis} plots.
% %%
% Furthermore, calculations based on a prior data selection can be redone within the report.
% The web-based approach allows us to quickly exchange results with our collaborating epidemiological experts.
% %%
% They can use the technique without installing any software.
% %%
% Exchanging the prototype becomes as easy as exchanging a hyperlink.
% %%
% The prototype is available at \url{ivapp15.dnsalias.com}.

@@

% \begin{figure*}[p!]

@@

%%Interactive scatter plot visualization of all classification results.

@@

no group (a), subdivision into \emph{male} and \emph{female} subjects (b) and quartiles of \emph{Body Mass Index} (\emph{BMI}) (c). % and quartiles of \emph{body size} (d).

@@

% We carry out the analysis using different subject groups:
% \begin{enumerate}
% 	%%
% 	\item All subjects,
% 	%%
% 	\item subdivision into \emph{male} and \emph{female},
% 	%%
% 	\item subdivision by \emph{Body Mass Index quantiles} ($BMI = \frac{m}{l^2}$ where $m$ is the \emph{body mass} in kilogram and $l$ is the \emph{body size} in meter), yielding the groups (17, 24.7] (24.7, 27.4] (27.4, 30.5] (30.5, 48],
% 	%%
% 	\item subdivision by \emph{size quantiles}, yielding the groups (139, 164], (164, 171], (171, 177], (177, 202].
% 	%%
% \end{enumerate}

@@

%The second plot shows all categorical variables transformed into dichotomous dummy variables. % (recall subsection~\ref{sec:C5.0}).

@@

%

@@

%The dichotomous variable whether the subject has felt \emph{pain in the last seven days} can also be predicted for this group using the same features.
%The resulting tree consists of $8$ rules and has an error of $0.35$.

@@

% \subsection{Size Groups}
% %%
% Many previously described results are influenced by \emph{subject size}.
% %%
% Differences between \emph{male} and \emph{female} subjects can be explained by the average \emph{body size} difference.
% %%
% For example, large subjects are already characterized by their rather straight spine.
% %
% The question is whether the inter-group spine variability parameter is sufficient for predicting other variables.
% %%
% Dividing subjects into \emph{body size} groups potentially highlights classifications not influenced by \emph{body size}.
% %%
% \\\\
% \noindent \textbf{Large Decision Trees.}
% Back pain-associated variables can be explained for various \emph{size} groups, but we could not extract universal rules.
% %%
% \emph{Radiating back pain} could be described with an error of $0.39$ using 23 rules for subjects between \emph{$171-177~cm$ body size} with \emph{torsion} and \emph{mean curvature}.
% %%
% For subjects sized \emph{$177-202~cm$} the error increases to $0.47$ using 20 decision rules.
% %%
% There are several decision trees for laboratory values, e.g. \emph{alanine aminotransferase} value (relevant for diagnosis of liver or gallbladder illness) in the blood can be described with an error of $0.4$ (\emph{$139-164~cm$}) to $0.36$ (\emph{$164 - 171~cm$}).
% %%
% Similar values can be observed for \emph{cholesterol} or \emph{age}.
% %%
% Due to the large decision trees, these results are not usable and impose overfitting to the data.
% %%
% \\\\
% \noindent \textbf{Small Decision Trees.}
% %%
% The dummy variables show several variables described using only one decision rule with an error between $0.42$ and $0.47$.
% %%
% Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
% %%
% These variables include \emph{nutrition}, \emph{thyroid disorder} and \emph{social problems induced by back pain}.
% %%

@@

% \noindent The presented results indicate that the image-based variables \emph{torsion}, \emph{curvature} and \emph{curvature angle} are not sufficient to characterize pathological changes in the lumbar spine in the given quality.
% %%
% They are associated with subject \emph{gender} due to their \emph{body size} difference using \emph{mean curvature} and \emph{mean curvature angle}.
% %%
% This is due to the higher curvature of smaller subjects and the significant difference of average \emph{body size} between male and female.
% %%
% \emph{Gender} was detected well in the analysis of all subjects as well as subjects divided by \emph{BMI}.
% Subjects grouped using \emph{body size} have a similar spine shape and a lower variance.
% %%
% The remaining information within these groups is not sufficient to characterize back pain-related variables.
% %%
% Associations with certain \emph{nutrition} or \emph{psychological} variables can be observed with a comparatively high error for some \emph{body size} groups.
% %%
% Another observation during the analysis is that large decision trees achieve low error scores.
% Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.

@@

% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
% Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
% %%
% This can be achieved by creating a describe several (dichotomous) target values with a decision tree constructed from all available data
% We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
% %%
% This allows to assess both the discriminative power of the data set regarding the target variables, and as the most important variables for the classification.

@@

%Another interesting metric is the spine canal thickness over the whole spine, capturing early signs of herniated disc disease.

@@

% \vfill
\bibliographystyle{style/apalike}
{\small
\bibliography{bibliography}}

%\section*{\uppercase{Appendix}}
%\noindent The IVA tool and all shown plots presented in this paper are available as supplementary material under the following link: \url{ivapp15.dnsalias.com}.

>>>>

@@

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1}, Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1} Department of Simulation and Graphics, University of Magdeburg, Germany}
\affiliation{\sup{2} Institute for Community Medicine, University of Greifswald, Germany}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald, Germany}
\email{klemm@isg.cs.uni-magdeburg.de}
}

@@

\abstract{ 
Epidemiology aims to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess potential relations between image-derived variables of the lumbar spine with lower back pain in a cross-sectional study.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as spine canal curvature and torsion.
%%
We analyze mutual dependencies among shape-describing variables and non-image variables, e.g., pain indicators.
%%
Therefore, we automatically train a decision tree classifier for each non-image variable.
%%
We provide an IVA technique to compare classifiers with a \emph{decision tree quality plot}.
%%
As a first result, we conclude that image-based variables are only sufficient to describe lifestyle factors within the data.
%%
A correlation between lumbar spine shape and lower back pain could not be found with the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, relations between non-image variables were successfully detected and described.
}

@@

\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
Large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
%%
This information is used to determine risk factors for diseases, inform people about healthier lifestyles or to support the diagnosis of widespread diseases.
%%
Cohort studies are an instrument of epidemiological research.
%%
We analyze back pain, one of the most frequent complaints in the Western civilization.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar spine attributes is still missing.
%%

We present an analysis of the image-derived data from a cohort study lumbar spine dataset.
%%
We extract possible associations between spine shape and back pain characteristics.
%%
For this purpose, we combine classification algorithms with data visualization techniques.
%%
Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
%%
We focus on highlighting new correlations and trigger \emph{hypotheses generation}, rather than statistically validate complex epidemiological correlations.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification of lumbar spine shape properties potentially associated with back pain,
%%
\item The detection of associations between image-based, socio-demographic and medical variables for \emph{hypotheses generation},
%%
\item The identification of the most important variables via classification methods using a novel \emph{decision tree quality plot}.
%%
\end{itemize}
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
(1) physicians formulate hypotheses based on observations made in their clinical practice; (2) epidemiologists compile a list of variables depicting a hypothesis to prove it; (3) statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated condition.
%%
Mutually dependent variables make this analysis challenging.
%%
For example, the incidence rate of many diseases, such as different cancer types increases with age \cite{Fletcher}.
%%------------------------------
\paragraph{Challenges of Epidemiological Data Analysis.}
%%
We focus on epidemiological large scale cohort study data.
%%
The collected data can be analyzed regarding many diseases and conditions.
%%
Epidemiological data are heterogenous and incomplete.
%%
For example, data about a disease treatment only affects subjects suffering from this condition.
%%
Statistical analysis has to take missing data into account.
%%
Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self-reported questionnaires or genetic examinations.
%%
This yields a heterogenous information space.
%%
Information reduction techniques are necessary to compare these data.
%%
For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\noindent Our work is based on a data set compiled by Epidemiologists with a wide range of variables possibly correlating with lumbar back pain, comprising non-image and image-derived data for 3,234 subjects.

@@

@@

All categorical variables are converted into binary \emph{dummy variables}, indicating the presence or absence of a categorical variable manifestation.

@@

@@

@@

@@

@@

@@

The detection fails for several subjects due to imaging artifacts or strongly deformed spines, yielding models for $2,540$ out of $3,234$ subjects.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

These observations brought us to the decision to incorporate classification techniques to assess the influence of the image-derived variables.
%%
\section{\uppercase{Interactive Decision Tree Quality Plot}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification to obtain a complex rule set on how combinations of the image-variables explain non-image variables.
%%
Decision trees are used to create predictive models.
%%
These models are built w.r.t. all input variables and capture more complex relationships than correlation coefficients.
%%
Leafs of a decision tree represent class labels, branches represent variable conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
Too many branches impose overfitting to the data \cite{DecisionTree}.
%%
\\\\
\noindent \textbf{Creating Decision Trees.}
We use the C4.5 algorithm, which builds decision trees based on information entropy.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
%%

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

no group (a), subdivision into \emph{male} and \emph{female} subjects (b) and quartiles of \emph{Body Mass Index} (\emph{BMI}) (c).

@@

@@

@@

@@

@@

@@

@@

@@

@@

\bibliographystyle{style/apalike}
{\small
\bibliography{bibliography}}
